---
title: "Advanced features"
author: Lucas da Cunha Godoy
format:
  html:
    html-math-method: mathjax
bibliography: [notes.bib, zim.bib]
vignette: >
  %\VignetteIndexEntry{Advanced features}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

## TL; DR {.unnumbered}

This document demonstrates some of the functionalities of the `drmr` R package
for fitting Dynamic Range and Species Distribution Models (DRMs and SDMs,
respectively) using pre-compiled `Stan` models.


We will show how to:

1. Visualize the estimated relationship between demographic processes and the
   environment
1. Visualize and assess out of sample predictions
1. Visualize the estimated age-specific densities


## Setup

The code below loads the packages necessary to reproduce the examples in this
document.

```{r}
#| label: setup

library(drmr)
library(sf) ## "mapping"
library(ggplot2) ## graphs
library(bayesplot) ## and more graphs
library(dplyr)
```

## Data {#sec-data}

For this example, we'll use the package's built-in _Summer flounder_ dataset,
which resembles the data analyzed in @fredston2025dynamic. We load it by
running:

```{r}
#| label: data-load

## loads the data
data(sum_fl)

## computing density
sum_fl <- sum_fl |>
  mutate(dens = 100 * y / area_km2,
         .before = y)
```

A quick exploration of the data reveals the following:

* Surface salinity (ssalin) and bottom salinity (bsalin) have missing values
  (around 30% missing for each).

* At least 25% of the density (`dens`) values are zero.

Finally, we split the data, reserving the last five years for evaluating
predictions:

```{r}
#| label: data-split

## 5 years-ahead predictions
first_year_forecast <- max(sum_fl$year) - 4

first_id_forecast <-
  first_year_forecast - min(sum_fl$year) + 1

years_all <- order(unique(sum_fl$year))
years_train <- years_all[years_all < first_id_forecast]
years_test <- years_all[years_all >= first_id_forecast]

## splitting data
dat_test <- sum_fl |>
  filter(year >= first_year_forecast)

dat_train <- sum_fl |>
  filter(year < first_year_forecast)
```


When fitting models with explanatory variables, it's good practice to
standardize them by centering (subtracting the mean) and scaling (dividing by
the standard deviation). This transformation to zero mean and unit variance
offers two main benefits: it typically improves the efficiency of Monte Carlo
Markov Chain (MCMC) sampling, and it makes comparing the magnitude of regression
coefficients easier for assessing relative variable importance. For applications
involving forecasting or prediction, it is crucial to calculate the mean and
standard deviation for standardization using only the training dataset. These
exact same values must then be applied to standardize the test dataset (or any
future data), simulating the real-world scenario where information about future
observations is unavailable during model fitting.


In that spirit, we center and scale some of our potential explanatory variables
as follows:

```{r}
#| label: center-scale

avgs <- c("stemp" = mean(dat_train$stemp),
          "btemp" = mean(dat_train$btemp),
          "depth" = mean(dat_train$depth),
          "n_hauls" = mean(dat_train$n_hauls),
          "lat" = mean(dat_train$lat),
          "lon" = mean(dat_train$lon))

min_year <- dat_train$year |>
  min()

## centering covariates
dat_train <- dat_train |>
  mutate(c_stemp = stemp - avgs["stemp"],
         c_btemp = btemp - avgs["btemp"],
         c_hauls = n_hauls - avgs["n_hauls"],
         c_lat   = lat - avgs["lat"],
         c_lon   = lon - avgs["lon"],
         time  = year - min_year)

dat_test <- dat_test |>
  mutate(c_stemp = stemp - avgs["stemp"],
         c_btemp = btemp - avgs["btemp"],
         c_hauls = n_hauls - avgs["n_hauls"],
         c_lat   = lat - avgs["lat"],
         c_lon   = lon - avgs["lon"],
         time  = year - min_year)
```


> For an extensive list of the variables present in this dataset run: `?sum_fl`.


## Fitting models

The models fitted here will explore some of the interesting features available
in the package. Auxiliary information on age-specific mortality rates and an
adjacency matrix will be used to leverage the DRMs. The adjacency matrix simply
indicate which patches share boarders with each other. That information may be
used either for spatial random effects or to establish a movement component in
the latent underlying population dynamics.


The mortality rates associated with the _Summer flounder_ dataset are shipped
with the package. The mortality rates are stored in a matrix having the number
of columns corresponding to the number of timepoints in our dataset, and number
of columns corresponding to the total number of age-groups observed for the
species. To load that matrix in our `R` session, we run:

```{r}
#| label: mortality_rates

fmat <-
  system.file("fmat.rds", package = "drmr") |>
  readRDS()

## splitting between train and test for model assessment
f_train <- fmat[, years_train]
f_test  <- fmat[, years_test]
```

Now, we load a shapefile associated with the patches used in the dataset to
construct the adjacency matrix:
```{r}
#| label: adj_mat

## loading map
shp_sum_fl <- system.file("maps/sum_fl.shp", package = "drmr") |>
  st_read()

## constructing adjacency matrix
adj_mat <- gen_adj(st_buffer(st_geometry(shp_sum_fl),
                             dist = 2500))
```

> The `shp_sum_fl` has one row for each patch. It is important that this dataset
> is ordered by the patch id, which should correspond to the patch id in the
> `sum_fl` dataset as well.


Finally, let us fit a model. The model fitted by the `fit_drm` call below takes
into account the following:

1. Effort: The probability of absence (i.e., probability of observing a zero
   density) is a function of the number of hauls, which serves as a proxy for
   effort.
   
1. Recruitment: We assume the relationship between recruitment and sea surface
   temperature is quadratic. In addition, we also include an AR(1) term for
   recruitment (`ar_re = "rec"`).

1. Survival: We estimate a baseline survival rate which is shared among all
   patches. The survival rates vary over time according to the external
   information provided by `f_train`.

1. Movement: We assume individuals from ages 3 to 10 may move between patches
   from one year to another.

1. Initialization: We use the survival rates to initialize the population
   dynamics (see the initialization vignette by running `vignette("init",
   "drmr")`.

```{r}
#| label: drm_rec

algo_args <- list(parallel_chains = 2,
                  chains = 2,
                  iter_sampling = 200,
                  iter_warmup = 500,
                  show_messages = FALSE,
                  show_exceptions = FALSE)

drm_rec <-
  fit_drm(.data = dat_train,
          y_col = "dens", ## response variable: density
          time_col = "year", ## vector of time points
          site_col = "patch",
          family = "gamma",
          seed = 2026,
          formula_zero = ~ 1 + c_hauls,
          formula_rec = ~ 1 + c_stemp + I(c_stemp * c_stemp),
          formula_surv = ~ 1,
          f_mort = f_train,
          n_ages = NROW(f_train),
          adj_mat = adj_mat, ## A matrix for movement routine
          ages_movement = c(0, 0,
                            rep(1, 12),
                            0, 0), ## ages allowed to move
          .toggles = list(ar_re = "rec",
                          movement = 1,
                          est_surv = 1,
                          est_init = 0,
                          minit = 1),
          algo_args = algo_args)
```

The default algorithm for posterior sampling is `Stan`'s NUTS. You can control
its settings by passing a named list to the `algo_args` argument in
`fit_drm`. Other available algorithms for inference are covered in a deficated
vignette. For further details, see `vignette("algos", "drmr")`.


Now, we also fit an alternative model with only one difference: it assumes the
survival rates vary according to the sea bottom temperature, while recruitment
is constant across patches (but varies over time). We will do it using the
`update` method. In that case, all the other arguments passed to `fit_drm` are
kept the same as in the last call.

```{r}
#| label: drm_srv

drm_srv <-
  update(drm_rec,
         formula_rec = ~ 1,
         formula_surv = ~ 1 + c_btemp + I(c_btemp * c_btemp))
```

## Environment and demographic processes

We can easily visualize the relationship between the environmental variables and
specific demographic processes using the `marg` function. The output of that
function can be plotted by simply calling the `plot` function. Below, we plot
both the relationship between recruitment and sea surface temperature for the
first model we fit, and between survival and sea bottom temperature for the
second model fitted.

```{r}
#| label: env_dem

marg(drm_rec, ## model
     process = "rec", ## demographic process
     variable = "c_stemp", ## environmental variable
     prob = .8) |> ## mass of the creible interval
  plot()

marg(drm_srv, ## model
     process = "surv", ## demographic process
     variable = "c_btemp", ## environmental variable
     prob = .8) |> 
  plot()
```

These figures display the centered environmental variables on the x-axis. We can
further customize the plots to have the environmental variable on the original
scale and, moreover, visualize different credible intervals at once. Below, we
show how to achieve that for the survival plot:

```{r}
#| label: custom_envdem

multiple_cis <-
  marg(drm_srv, process = "surv", variable = "c_btemp",
       summary = FALSE,
       n_pts = 200) |>
  lapply(c(0.6, 0.8, 0.9, 0.99),
         \(pr, .dt) {
           alpha <- (1 - pr) / 2
           probs <- c(alpha, 1 - alpha)
           .dt |>
             group_by(c_btemp) |>
             summarise(srv = median(survival),
                       low = quantile(survival, probs = probs[1]),
                       upp = quantile(survival, probs = probs[2])) |>
             ungroup() |>
             mutate(prob_mass = pr)
         }, .dt = _) |>
  bind_rows() |>
  mutate(c_btemp = c_btemp + avgs["btemp"])

ggplot(data = multiple_cis,
       aes(x = c_btemp,
           group = factor(prob_mass,
                          levels = rev(levels(factor(prob_mass)))))) +
  geom_rug(data = dat_train,
           aes(x = btemp),
           inherit.aes = FALSE,
           alpha = 0.3,
           length = unit(0.02, "npc"),
           colour = "grey40") +
  geom_ribbon(aes(ymin = low, ymax = upp,
                  fill = factor(prob_mass,
                                levels = rev(levels(factor(prob_mass))))),
              colour = NA) +
  geom_line(aes(y = srv), linewidth = 0.6, colour = "white") +
  scale_fill_manual(
    values = c("0.99" = "#d1e5f0",
               "0.9"  = "#6baed6",
               "0.8"  = "#2171b5",
               "0.6"  = "#08306b"),
    labels = \(x) paste0(as.numeric(x) * 100, "%")
  ) +
  labs(x    = "Sea bottom temperature",
       y    = "Survival",
       fill = "Credible interval") +
  theme_minimal(base_size = 13) +
  theme(
      legend.position  = "inside",
      legend.position.inside = c(.2, .75),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(colour = "grey90")
  )
```

From the graph, on may infer that the temperature maximizes the survival rates
is around 15 degrees Celsius.


## Comparing out-of-sample predictions

The fitted values (or in-sample predictions) are obtained from the draws of the
posterior predictive distribution. In general, we will be interested in
summarizing those draws for visualization. The code chunk below shows how to use
the `fitted` and `summary` methods to achieve that goal.

```{r}
#| label: fitted

fitted_rec <- fitted(drm_rec) |>
  summary() |>
  mutate(model = "rec", .before = 1) ## creating a column that identifies the ##
                                     ## model

fitted_srv <- fitted(drm_srv) |>
  summary() |>
  mutate(model = "srv", .before = 1) ## creating a column that identifies the
                                     ## model
```

The out-of-sample predictions are obtained just as easily:
```{r}
#| label: preds

predicted_rec <- predict(drm_rec,
                         new_data = dat_test,
                         past_data = filter(dat_train,
                                            year == max(year)),
                         seed = 2026,
                         f_test = f_test) |>
  summary() |>
  mutate(model = "rec", .before = 1) ## creating a column that identifies the ##
                                     ## model

predicted_srv <- predict(drm_srv,
                         new_data = dat_test,
                         past_data = filter(dat_train,
                                            year == max(year)),
                         seed = 2026,
                         f_test = f_test) |>
  summary() |>
  mutate(model = "srv", .before = 1) ## creating a column that identifies the
                                     ## model
```


We can now combine the in- and out-of-sample predictions to visualize the
results. Moreover, we will join the predictions outputs to the original data to
enhance the visualizations.

```{r}
#| label: bind_and_join

combined_sfl <-
  bind_rows(fitted_rec, predicted_rec,
            fitted_srv, predicted_srv) |>
  mutate(patch = as.integer(patch)) |>
  left_join(sum_fl, by = c("patch", "year"))


ggplot(data = combined_sfl,
       aes(x = year)) +
  geom_ribbon(aes(ymin = q5, ymax = q95),
              alpha = .5) +
  geom_line(aes(y = q50)) +
  geom_point(aes(y = dens), color = 2, pch = 19)+ 
  facet_grid(model ~ patch, scales = "free_y") +
  theme_bw()
```

We can objectively compare the predictions as well.

```{r}
#| label: rmse

combined_sfl <-
  combined_sfl |>
  mutate(type = ifelse(year >= first_year_forecast,
                       "out-of-sample",
                       "in-sample"))

combined_sfl |>
  mutate(mse = dens - q50) |>
  mutate(mse = mse * mse) |>
  group_by(type, model) |>
  summarise(rmse = sqrt(mean(mse))) |>
  knitr::kable()
```

As expected, in-sample predictions have lower root mean square error (RMSE) of
prediction when compared to their out-of-sample counterpart. In both cases, the
model that relates recruitment to the environment has lead to better predictive
performance (in terms of RMSE).

## Projections under unknown effort

Often, you may want to project the abundance (measured here as density) of a
species over time. However, since future sampling effort is unknown, it is
crucial to isolate this source of variation.


`drmr` can handle this by projecting the latent density. Latent density
represents the true, unobserved population at a given site. Because our models
assume that observed data captures only a subset of the total population at a
given site and time, the latent density acts as a proxy for true abundance and
will consistently exceed the observed abundance. Furthermore, because it
isolates the underlying biological process from observation error, the latent
density trajectory tends to be much "smoother".


You can compute the in- and out-of-sample latent abundances by passing the `type
= "latent"` argument to the `fitted` and `predict` functions:
```{r}
#| label: latent

fitted_rl <- fitted(drm_rec, type = "latent") |>
  summary() 


predicted_rl <- predict(drm_rec,
                        new_data = dat_test,
                        past_data = filter(dat_train,
                                           year == max(year)),
                        seed = 2026,
                        f_test = f_test,
                        type = "latent") |>
  summary()

combined_rl <- bind_rows(fitted_rl, predicted_rl)
```

Now, we can visualize those:
```{r}
#| label: viz_latent
combined_rl |>
  mutate(patch = as.integer(patch)) |>
  ## to allow for a comparison with reality
  left_join(sum_fl, by = c("patch", "year")) |>
  ggplot(data = _,
         aes(x = year)) +
  geom_ribbon(aes(ymin = q5, ymax = q95),
              alpha = .5) +
  geom_line(aes(y = q50)) +
  geom_point(aes(y = dens), color = 2, pch = 19)+ 
  facet_wrap(. ~ patch, scales = "free_y") +
  theme_bw()
```


## Visualizing age specific densities

Our package also allows for estimating the age-specific densities. To compute
those, one can simply run:
```{r}
#| label: edens

lambdas <- ages_edens(drm_rec) |>
  summary() |>
  mutate(patch = as.integer(patch))
```

Now, let's visualize the age-specific densities for all patches at a given year:
```{r}
#| label: edens_year

lambdas |>
  filter(year == 2010) |>  
  ggplot(data = _,
         aes(x = age)) +
  geom_segment(aes(xend = age, y = 0, yend = q50),
               color = "gray50") +
  geom_errorbar(aes(ymin = q5, ymax = q95)) +
  geom_point(aes(y = q50)) +
  facet_wrap(~ patch) +
  theme_bw()

```

The error bars represent the respective 90\% credible intervals. An equivalent
visualization for a single patch across time can be obtained as follows:
```{r}
#| label: edens_patch

lambdas |>
  filter(patch == 5) |>
  filter(year %in% seq(1985, 2010, by = 5)) |>
  ggplot(data = _,
         aes(x = age)) +
  geom_segment(aes(xend = age, y = 0, yend = q50),
               color = "gray50") +
  geom_errorbar(aes(ymin = q5, ymax = q95)) +
  geom_point(aes(y = q50)) +
  facet_wrap(~ year) +
  theme_bw()

```



## References
