---
title: "Algorithms"
author: Lucas da Cunha Godoy
format:
  html:
    html-math-method: mathjax
bibliography: [notes.bib, zim.bib]
vignette: >
  %\VignetteIndexEntry{Algorithms}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

## TL; DR {.unnumbered}

This document demonstrates how to modify the algorithm used for inference in our
models.

## Setup

The setup and data-processing for this vignette is the same as in the "Advanced
features" vignette. For further details, see `vignette("advanced-features",
"drmr")`. Unlike the examples in that vignette, we will not split the data into
train and test, but rather just look at parameters' estimates obtained through
different algorithms.

```{r}
#| label: setup

library(drmr)
library(sf) ## "mapping"
library(ggplot2) ## graphs
library(bayesplot) ## and more graphs
library(dplyr)

## loads the data
data(sum_fl)

## computing density
sum_fl <- sum_fl |>
  mutate(dens = 100 * y / area_km2,
         .before = y)

avgs <- c("stemp" = mean(sum_fl$stemp),
          "btemp" = mean(sum_fl$btemp),
          "depth" = mean(sum_fl$depth),
          "n_hauls" = mean(sum_fl$n_hauls),
          "lat" = mean(sum_fl$lat),
          "lon" = mean(sum_fl$lon))

min_year <- sum_fl$year |>
  min()

## centering covariates
sum_fl <- sum_fl |>
  mutate(c_stemp = stemp - avgs["stemp"],
         c_btemp = btemp - avgs["btemp"],
         c_hauls = n_hauls - avgs["n_hauls"],
         c_lat   = lat - avgs["lat"],
         c_lon   = lon - avgs["lon"],
         time  = year - min_year)

## mortality rates
fmat <-
  system.file("fmat.rds", package = "drmr") |>
  readRDS()

shp_sum_fl <- system.file("maps/sum_fl.shp", package = "drmr") |>
  st_read()

## constructing adjacency matrix
adj_mat <- gen_adj(st_buffer(st_geometry(shp_sum_fl),
                             dist = 2500))
```


> For an extensive list of the variables present in this dataset run: `?sum_fl`.


## Fitting a model using different algorithms


We will start with the default algorithm (NUTS). By default, it runs 4 chains
sequentially, with a warmup period od 1000 samples and, subsequently, 1000
samples per chain. Here, we are running two chains in parallel, with a warmup of
500 and drawing 500 samples.

```{r}
#| label: nuts

nuts_args <- list(parallel_chains = 2,
                  chains = 2,
                  iter_sampling = 500,
                  iter_warmup = 500,
                  show_messages = FALSE,
                  show_exceptions = FALSE)

drm_nuts <-
  fit_drm(.data = sum_fl,
          y_col = "dens", ## response variable: density
          time_col = "year", ## vector of time points
          site_col = "patch",
          family = "gamma",
          seed = 2026,
          formula_zero = ~ 1 + c_hauls,
          formula_rec = ~ 1 + c_stemp + I(c_stemp * c_stemp),
          formula_surv = ~ 1,
          f_mort = fmat[, -1],
          n_ages = NROW(fmat),
          adj_mat = adj_mat, ## A matrix for movement routine
          ages_movement = c(0, 0,
                            rep(1, 12),
                            0, 0), ## ages allowed to move
          .toggles = list(ar_re = "rec",
                          movement = 1,
                          est_surv = 1,
                          est_init = 0,
                          minit = 1),
          algo_args = nuts_args)
```


Next, we obtain samples from the _variational_ posterior using Stan's Automatic
Differentiation Variational Inference (ADVI) algorithm (for for details see
[this link](https://mc-stan.org/docs/cmdstan-guide/variational_config.html)).


```{r}
#| label: advi

advi_args <- list(show_messages = FALSE,
                  show_exceptions = FALSE,
                  iter = 10^5, ## maximum number of iterations (optimization)
                  draws = 4000) ## number of samples from the posterior

drm_advi <-
  fit_drm(.data = sum_fl,
          y_col = "dens", ## response variable: density
          time_col = "year", ## vector of time points
          site_col = "patch",
          family = "gamma",
          seed = 2026,
          formula_zero = ~ 1 + c_hauls,
          formula_rec = ~ 1 + c_stemp + I(c_stemp * c_stemp),
          formula_surv = ~ 1,
          f_mort = fmat[, -1],
          n_ages = NROW(fmat),
          adj_mat = adj_mat, ## A matrix for movement routine
          ages_movement = c(0, 0,
                            rep(1, 12),
                            0, 0), ## ages allowed to move
          .toggles = list(ar_re = "rec",
                          movement = 1,
                          est_surv = 1,
                          est_init = 0,
                          minit = 1),
          algo_args = advi_args,
          algorithm = "vb") ## vb stands for variational Bayes
```


Another algorithm option is the
[Pathfinder](https://mc-stan.org/docs/cmdstan-guide/pathfinder_config.html). Pathfinder
also relies on variational inference. It is usually better than ADVI, especially
when the posteriors (in the unconstrained space) are not unimodal and
bell-shaped. We will demonstrate how the inference algorithm can be updated
using the `update` method:


```{r}
#| label: pathfinder

path_args <- list(show_messages = FALSE,
                  show_exceptions = FALSE,
                  max_lbfgs_iters = 10^4)

drm_path <-
  update(drm_advi,
         algo_args = path_args,
         algorithm = "pathfinder")
```


It is also possible to obtain samples from a Laplace approximation of the
posterior as follows:

```{r}
#| label: laplace

lapl_args <- list(show_messages = FALSE,
                  show_exceptions = FALSE)

drm_lapl <-
  update(drm_path,
         algo_args = lapl_args,
         algorithm = "laplace")
```


The code below computes the parameter estimates for each of the methods, and
then makes a graph to compare them.
```{r}
#| label: parest

bind_rows(
    mutate(summary(drm_nuts)$estimates, algo = "nuts"),
    mutate(summary(drm_advi)$estimates, algo = "advi"),
    mutate(summary(drm_path)$estimates, algo = "path"),
    mutate(summary(drm_lapl)$estimates, algo = "lapl")
) |>
  ggplot(data = _,
         aes(x = q50,
             y = algo)) +
  geom_linerange(aes(xmin = q5, xmax = q95)) +
  geom_point() +
  facet_wrap(~ variable, scales = "free_x") +
  theme_bw()
```

In general, variational inference methods are known for underestimating the
uncertainty around the parameter estimates. We can also look at the estimated
relationship between environment and recruitment for each of those methods:
```{r}
#| label: envrec
#| layout-ncol: 2
#| layout-nrow: 2

marg(drm_nuts, "rec", "c_stemp") |>
  plot() +
  labs(title = "nuts")

marg(drm_path, "rec", "c_stemp") |>
  plot() +
  labs(title = "path")

marg(drm_advi, "rec", "c_stemp") |>
  plot() +
  labs(title = "advi")

marg(drm_lapl, "rec", "c_stemp") |>
  plot() +
  labs(title = "lapl")
```


## References
