---
title: "Examples"
author: Lucas da Cunha Godoy
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  eval,
  comment = "#>"
)
```

---

## TL; DR {.unnumbered}

This document demonstrates the functionalities of the `drmr` R package for
fitting Dynamic Range and Species Distribution Models (DRMs and SDMs,
respectively) using pre-compiled `Stan` models.


We'll explore increasingly complex models, showcasing the package's capabilities
with the help of five additional `R` packages:

* `sf`: for spatial data manipulation and visualization
* `ggplot2`: for creating graphs
* `bayesplot`: for visualizing MCMC outputs
* `dplyr`: for data wrangling
* `skimr`: for quick descriptive statistics

## Installing the `drmr` package {#sec-drmr}

The `drmr` package provides pre-compiled `cmdstanr` [@gabry2024cmdstanr] models
powered by the `instantiate` package [@landau2024instantiate]. These
pre-compiled models allow us to make inferences about the DRM (and SDM) using
`cmdstan` algorithms, such as the HMC NUTS Sampler.


The package is not on CRAN yet. To install the version hosted on GitHub, run:
```{r}
#| eval: false

remotes::install_github("pinskylab/drmr")
```

Finally, the code below loads the packages necessary to reproduce the examples
in this document.

```{r}
#| label: setup

library(drmr)
library(sf) ## "mapping"
library(ggplot2) ## graphs
library(bayesplot) ## and more graphs
library(dplyr)
```

## Data {#sec-data}

This example uses Summer Flounder data from Alexa's project.  We begin by
loading the dataset and calculating a density variable (raw counts divided by
patch area):

```{r}
#| label: data-load

## loads the data
data(sum_fl)

## computing density
sum_fl <- sum_fl |>
  mutate(dens = y / area_km2,
         .before = y)
```

A quick exploration of the data in @tbl-summary reveals the following:

* Surface salinity (ssalin) and bottom salinity (bsalin) have missing values
  (around 30% missing for each).
* At least 25% of the density (`dens`) values are zero.

```{r}
#| label: tbl-summary
#| tbl-cap: "Summary statistics of all the variables in the dataset."

## Density is being multiplied by 10000 because the non-negative densities are
## too small
skimr::skim_without_charts(mutate(sum_fl, dens = 10000 * dens))
```

Finally, we split the data, reserving the last five years for evaluating
predictions:

```{r}
#| label: data-split

## 5 years-ahead predictions
first_year_forecast <- max(sum_fl$year) - 5

first_id_forecast <-
  first_year_forecast - min(sum_fl$year) + 1

## splitting data
dat_test <- sum_fl |>
  filter(year >= first_year_forecast)

dat_train <- sum_fl |>
  filter(year < first_year_forecast)
```


## Loading pre-compiled models


To load the pre-compiled SDM described in @sec-sdm, run:
```{r}
#| label: load-sdm

sdm_compiled <-
  instantiate::stan_package_model(name = "sdm",
                                  package = "drmr")
```


Similarly, to load the pre-compiled DRM, we run:
```{r}
#| label: load-drm

drm_compiled <-
  instantiate::stan_package_model(name = "drm",
                                  package = "drmr")
```


These models also have associated pre-compiled `forecast` objects. They can be
loaded as follows:
```{r}
#| label: load-forecasts

forecast_sdm <-
  instantiate::stan_package_model(name = "forecast_sdm",
                                  package = "drmr")

forecast_drm <-
  instantiate::stan_package_model(name = "forecast",
                                  package = "drmr")
```

Lastly, to calculate density-weighted centroids, the following pre-compiled
object is available:
```{r}
#| label: load-centroids

centroids_compiled <-
  instantiate::stan_package_model(name = "centroid",
                                  package = "drmr")
```


All of the pre-compiled models are of class `CmdStanModel` and they work just as
any other compiled `cmdstanr`
[model](https://mc-stan.org/cmdstanr/reference/CmdStanModel.html#model-fitting).


## Preparing data & fitting models


Currently, in addition to the pre-compiled models, the `drmr` package has three
functions. They are the following:

* `make_data`: creates a `list` that will be used as the input for a `drm` model
  (see `?drmr::make_data` for all the function arguments).

* `default_priors` is a function that takes no arguments and return the default
   hyperparameters for the priors from the model;

* `default_toggles` is another function taking no arguments. This one return the
   default "toggles" for the `drm` the model. The toggles turn features of the
   model (e.g., movement) on and off.


Let us start fitting the simplest DRM available in the package. The first step,
is always to use the `make_data` function to create the input for the `cmdstan`
model. This function takes a set of inputs, and turn them into the appropriate
format for our code. Below, we only provide three inputs for the function:

* `y` the vector of densities

* `time` the vector of time points

* `site` the vector of patches/sites.

```{r}
#| label: simplest_makedata

data <-
  make_data(y = dat_train$dens, ## response variable: density
            time = dat_train$year, ## vector of time points
            site = dat_train$patch, ## vector of patches
            n_ages = 8, ## number of age groups
            .toggles = list(likelihood = 1))

## checking the elements of the list:
class(data)
names(data)
```

The output of `names(data)` displays the "names" of the elements of the output
list from the `make_data`. Some of these elements are inferred from `y`, `time`,
and `site` and others take default values. For example:

* `n_ages` is the total number of (unobserved) ages $A$. If not provided,
  `n_ages` is taken to be 1.

* `f` represents $f_{a, t}$ from @eq-surv. Its default value is $f_{a, t} = 0$ for
  every age and time point.

* `m` is the mortality rate $m$ in @eq-surv. Its default value is 0.25 [^2].

* `X_t` is a matrix such that its $i$-th row is as in $\mathbf{x}^{(1)}_{t, i}$
  in @eq-linpreds. While `K_t` represents the number of columns in `X_t`. If no
  input is given, `X_t` has a single column and $N = T \cdot P$ rows.
  
* `X_r` is a matrix such that its $i$-th row is as in $\mathbf{x}^{(r)}_{t, i}$
  in @eq-lmrec. While `K_r` represents the number of columns in `X_r`. If no
  input is given, `X_r` has a single column and $N = T \cdot P$ rows.
  
* `X_m` is a matrix such that its $i$-th row is as in $\mathbf{x}^{(m)}_{t, i}$
  in @eq-lmsurv. While `K_m` represents the number of columns in `X_m`. If no
  input is given, `X_m` is "empty".
  
* `adj_mat` is the adjacency matrix used to encode movement as $\mathbf{A}$,
  while `age_at_maturity` is $a_0$ in @sec-movement.

* `selectivity_at_age` is a vector of length `n_ages` informing the selectivity
  at each age group.


[^2]: taken from Alexa's code.


The remaining elements are toggles and prior hyperparameters that will be
described later on. The prior hyperparameters start with `pr_`. 


The toggle called `likelihood` takes four possible values. When `likelihood = 0`
(default), the $g(\cdot)$ function in @eq-za_dens is taken to be a Log-Normal
density with the usual parametrization. When `likelihood = 1`, we still get a
Log-Lormal density, but parametrized in terms of $\mu$ and $\phi$. For
`likelihood` equal to 2 or 3, we get a Gamma or a Log-Logistic distribution,
respectively. The last two densities are also parametrized in terms of their
mean. For more details, see @ye2021comparisons.



Finally, below we obtain samples from the model associated with the data list we
just created.
```{r}
#| label: drm-1

drm_1 <- drm_compiled$sample(data = data,
                             iter_sampling = 400, ## number of samples after warmup
                             iter_warmup = 400, ## number of warmup samples
                             seed = 2025,
                             chains = 4,
                             parallel_chains = 4)
```


This model has only three parameters:

* `phi`: this parameter has no practical interpretation;

* In this simpler model, $\theta_{t, i} = \theta$ for every $i$ and $t$. Then,
  `coef_t = \mathrm{logit}(\theta)` is the logit of the probability of observing
  a zero density.

* In this simpler model, `coef_r` is the log of the overall recruitment per unit
  of area.


We can introduce a AR(1) temporal random effect for recruitment by turning the
`time_ar` toggle on and re-running the model. See below the code chunk below.

```{r}
#| label: drm-2

data <-
  make_data(y = dat_train$dens, ## response variable: density
            time = dat_train$year, ## vector of time points
            site = dat_train$patch, ## vector of patches
            n_ages = 8, ## number of age groups
            .toggles = list(likelihood = 1, time_ar = 1))

drm_2 <- drm_compiled$sample(data = data,
                             iter_sampling = 400, ## number of samples after
                                                  ## warmup
                             iter_warmup = 400, ## number of warmup samples
                             seed = 2025,
                             chains = 4,
                             parallel_chains = 4)
```

This model introduces two additional parameters: `tau` (the conditional SD
of the AR term), and `alpha` the temporal autocorrelation parameter.


Now, let us incorporate the fishing mortality into the model. We can load the
Summer Flounder fishing mortality rates as follows:

```{r}
#| label: fmat

fmat <-
  system.file("fmat.rds", package = "drmr") |>
  readRDS()

years_all <- seq_len(NCOL(fmat))
years_train <- years_all[years_all < first_id_forecast]
years_test  <- years_all[years_all >= first_id_forecast]

f_train <- fmat[, years_train]
f_test  <- fmat[, years_test]
```

To fit a model taking this information into consideration, we just provide the
`fmat` object to `make_data` and refit the model. 

```{r}
#| label: drm-3

data <-
  make_data(y = dat_train$dens, ## response variable: density
            time = dat_train$year, ## vector of time points
            site = dat_train$patch, ## vector of patches
            f_mort = f_train,
            n_ages = NROW(f_train), ## number of age groups
            .toggles = list(likelihood = 1, time_ar = 1))

drm_3 <- drm_compiled$sample(data = data,
                             iter_sampling = 400, ## number of samples after
                                                  ## warmup
                             iter_warmup = 400,   ## number of warmup samples
                             seed = 2025,
                             chains = 4,
                             parallel_chains = 4)
```


Now, let us introduce movement into the model. First, let us load the map of the
patches to be able to calculate the adjacency matrix $\mathbf{A}$ from
@sec-movement.

```{r}
#| label: fig-map-adj
#| fig-cap: "Patches considered in this example."

map_name <- system.file("maps/sum_fl.shp", package = "drmr")

polygons <- st_read(map_name)

ggplot() +
  geom_sf(data = polygons,
          color = 1,
          fill = "gray70") +
  theme_bw()
```

Finally, the code below calculates the auxiliary quantities for the movement
matrix.

```{r}
#| label: calc-adj

adj_mat <- gen_adj(st_buffer(st_geometry(polygons),
                             dist = 2500))

## row-standardized matrix
adj_mat <-
  t(apply(adj_mat, 1, \(x) x / (sum(x))))
```

Below, we adjust the `make_data` function accordingly and fit the model with
movement.

```{r}
#| label: drm-4

data <-
  make_data(y = dat_train$dens, ## response variable: density
            time = dat_train$year, ## vector of time points
            site = dat_train$patch, ## vector of patches
            f_mort = f_train,
            n_ages = NROW(f_train), ## number of age groups
            adj_mat = adj_mat,
            age_at_maturity = 3, ## age at which fish are able to move
            .toggles = list(likelihood = 1, ## non-zero density pdf
                            time_ar = 1, ## ar(1) for recruitment
                            movement = 1)) ## turning movement on

drm_4 <- drm_compiled$sample(data = data,
                             iter_sampling = 400, ## number of samples after
                                                  ## warmup
                             iter_warmup = 400,   ## number of warmup samples
                             seed = 2025,
                             chains = 4,
                             parallel_chains = 4)
```


This model introduces the parameter `zeta` ($\zeta$) which represents the
probability of _not moving_ from time $t$ to time $t + 1$. The estimated
probability of not moving is $0.498$.



We can make the model more complex/flexible by incorporating environmental
drivers. We have two variables representing environmental drivers: sea bottom
temperature (`btemp`) and sea surface temperature (`stemp`). They can be
incorporated into the probability of observing zero density, recruitment, and/or
survival (or mortality).


The relationship between recruitment (or survival) and environmental factors we
can obtain are highlighted in @fig-rel. Before introducing these variables into
the models, it is a good practice to center them at 0 to improve the MCMC
efficiency. The code chunk below centers the covariates at 0.

```{r}
#| label: cov-centering

avgs <- c("stemp" = mean(dat_train$stemp),
          "btemp" = mean(dat_train$btemp))

min_year <- dat_train$year |> min()

dat_train <- dat_train |>
  mutate(stemp = stemp - avgs["stemp"],
         btemp = btemp - avgs["btemp"],
         time  = year - min_year)

dat_test <- dat_test |>
  mutate(stemp = stemp - avgs["stemp"],
         btemp = btemp - avgs["btemp"],
         time  = year - min_year)
```


```{r}
#| label: fig-rel
#| fig-cap: "Available relationship between recruitment/survival and environmental factors."
#| echo: false
#| fig-subcap: 
#|   - "Quadratic relationship 1"
#|   - "Quadratic relationship 2"
#|   - "log-linear relationship"
#| layout-ncol: 3

max_y <- function(x, .betas, type = "log", offset = 0) {
  .betas[1] <- .betas[1] - 2 * offset * .betas[2]
  max_x <- - .5  * .betas[1] / .betas[2]
  max_x2 <- max_x * max_x
  out <- as.numeric(cbind(1, max_x, max_x2) %*% .betas)
  if (type == "log") {
    return(exp(out))
  } else if (type == "logit") {
    return(expit(out))
  } else
    return(out)
}

t_max <- 2
b2 <- -0.25
b1 <- - 2 * t_max * b2
betas <- c(1, b1, b2)
my_df <- tibble(x = seq(from = 0, to = 20, length.out = 500)) |>
  mutate(x2 = x * x) |>
  mutate(ylin = as.numeric(cbind(1, x, x2) %*% betas)) |>
  mutate(ylog = exp(ylin))

## we can still get an analogous of "T_opt"
ggplot(data = my_df,
       aes(x = x, y = ylog)) +
  geom_line() +
  theme_bw()

t_max <- 11
b2 <- -0.05
b1 <- - 2 * t_max * b2
betas <- c(1, b1, b2)
my_df <- tibble(x = seq(from = 0, to = 20, length.out = 500)) |>
  mutate(x2 = x * x) |>
  mutate(ylin = as.numeric(cbind(1, x, x2) %*% betas)) |>
  mutate(ylog = exp(ylin))
## we can still get an analogous of "T_opt"
ggplot(data = my_df,
       aes(x = x, y = ylog)) +
  geom_line() +
  theme_bw()

betas <- c(.1, .25)
my_df <- tibble(x = seq(from = 0, to = 20, length.out = 500)) |>
  mutate(ylin = as.numeric(cbind(1, log(x)) %*% betas)) |>
  mutate(ylog = exp(ylin))

ggplot(data = my_df,
       aes(x = x, y = ylog)) +
  geom_line() +
  theme_bw()
```


A quadratic relationship between recruitment and `stemp`, a log-linear
relationship between survival and `btemp`, and a non-linear function between the
log-odds of observing a zero density and those two variables can be obtained as
in the code chunk below.

```{r}
#| label: building-x

##--- formulas (will be helpful for forecasting as well) ----

## formula for the linear predictor of recruitmemt it defines a quadratic
## relationship between recruitment and surface temperature
form_r <- ~ 1 +
  stemp +
  I(stemp * stemp)

## formula for the linear predictor in the survival rates
form_m <- ~ 1 + btemp + I(btemp * btemp)

## formula for the linear predictor of observing a zero density
form_t <- ~ 1 +
  btemp +
  I(btemp * btemp) +
  stemp +
  I(stemp * stemp) +
  I(btemp * stemp)

##--- building "X matrices" ----

x_m <- model.matrix(form_m,
                    data = dat_train)
x_r <- model.matrix(form_r,
                    data = dat_train)
x_t <- model.matrix(form_t,
                    data = dat_train)
```


Finally, the code below adjusts the data object for a model with movement,
time-varying and env-dependent recruitment, env-dependent mortality, and
env-dependent prob of observing a "0 density".

```{r}
#| label: data-5

data <-
  make_data(y = dat_train$dens, ## response variable: density
            time = dat_train$year, ## vector of time points
            site = dat_train$patch, ## vector of patches
            f_mort = f_train,
            n_ages = NROW(f_train), ## number of age groups
            x_t = x_t,
            x_r = x_r,
            x_m = x_m,
            adj_mat = adj_mat,
            age_at_maturity = 3, ## age at which fish are able to move
            .toggles = list(likelihood = 1, ## non-zero density pdf
                            time_ar = 1, ## ar(1) for recruitment
                            qr_t = 1, ## QR parametrization for mortality coefs
                            est_mort = 1, ## turn env dependent mortality on
                            movement = 1), ## turning movement on
            .priors = list(pr_phi_a = 9, ## hyperpar for phi prior (phi ~ Gamma(a, b))
                           pr_phi_b = .5, ## hyperpar for phi prior
                           pr_coef_r_mu = rep(0, ncol(x_r)), ## hyperpar for coef_r
                           pr_coef_r_sd = c(10000, rep(1, ncol(x_r) - 1)))) ## hyperpar for coef_r
```


The code below fits the model associated to the data object above to the data
and introduce another feature: using the pathfinder [@zhang2022pathfinder]
algorithm for obtaining the initial values for the MCMC algorithm.

```{r}
#| label: drm-5

## initial values for mcmc
init_5 <- 
  drm_compiled$pathfinder(data = data,
                          seed = 2025,
                          num_paths = 4,
                          save_single_paths = TRUE,
                          psis_resample = FALSE)


drm_5 <- drm_compiled$sample(data = data,
                             iter_sampling = 400, ## number of samples after
                                                  ## warmup
                             iter_warmup = 400,   ## number of warmup samples
                             seed = 2025,
                             chains = 4,
                             init = init_5,
                             parallel_chains = 4)
```


We can also fit a Gamma (instead of log-normal) model to the data. It
suffices to modify the `likelihood` toggle in the data object and rerun the
code. See below.

```{r}
#| label: drm-6

## modifying only "likelihood" toggle
data$likelihood <- 2

## initial values for mcmc
init_6 <- 
  drm_compiled$pathfinder(data = data,
                          seed = 2025,
                          num_paths = 4,
                          save_single_paths = TRUE,
                          psis_resample = FALSE)

drm_6 <- drm_compiled$sample(data = data,
                             iter_sampling = 400, ## number of samples after
                                                  ## warmup
                             iter_warmup = 400,   ## number of warmup samples
                             seed = 2025,
                             chains = 4,
                             init = init_6,
                             parallel_chains = 4)
```


Lastly, let us fit a similar Gamma SDM to the data. The data object associated
with the `sdm` model is similar, but some elements need to be renamed [^3]

[^3]: I'm working on building a similar function to create the data for the SDM
    as well.

```{r}
#| label: sdm-1

## copy the data object from the last model
data_sdm <- data

## modify time (need to start at 1 for the SDM)
data_sdm$time <- dat_train$time + 1

## renaming elements
names(data_sdm)[names(data_sdm) == "X_t"] <- "Z"
names(data_sdm)[names(data_sdm) == "K_t"] <- "K_z"
names(data_sdm)[names(data_sdm) == "X_r"] <- "X"
names(data_sdm)[names(data_sdm) == "K_r"] <- "K_x"
names(data_sdm)[names(data_sdm) == "qr_r"] <- "qr_x"
names(data_sdm)[names(data_sdm) == "qr_t"] <- "qr_z"

init_sdm <-
  sdm_compiled$pathfinder(data = data_sdm,
                          seed = 2025,
                          num_paths = 4,
                          save_single_paths = TRUE,
                          psis_resample = FALSE)

sdm_1 <-
  sdm_compiled$sample(data = data_sdm,
                      iter_sampling = 400,
                      iter_warmup = 400,
                      seed = 2025,
                      chains = 4,
                      parallel_chains = 4,
                      init = init_sdm)
```


To compare how well each of these models adjust to the data, we can look at
their leave-one-out information criterion [LOOIC;
@vehtari2017practical]. According to this criteria, the best model (in terms of
goodness-of-fit) was the most complex DRM with a Gamma pdf for the non-zero part
of the zero-augmented model. The SDM appears as the third best model according
to this criteria.

```{r}
#| label: loo-compare

all_loos <-
  list("DRM 1" = drm_1$loo(),
       "DRM 2" = drm_2$loo(),
       "DRM 3" = drm_3$loo(),
       "DRM 4" = drm_4$loo(),
       "DRM 5" = drm_5$loo(),
       "DRM 6" = drm_6$loo(),
       "SDM 1" = sdm_1$loo())


loo::loo_compare(all_loos)
```

According to the LOOIC criteria, the model without the AR(1) term has a better
fit.


## Diagnosing the best models


In this Section, we will look into the convergence of the chains associated to
some of the model parameters for the best DRM and SDM models.


### DRM

@fig-drm6_1 depicts the traceplots (left) and overlay-ed (right) density plots
for the parameter $\phi$ and the two parameters (`sigma_r` and $\rho$)
associated to the AR(1) random effect used to introduce temporal dependence to
recruitment. In general, the chains seem to mix well and the estimated posterior
densities associated to different chains are fairly similar, which is a good
indication of convergence.

```{r}
#| label: fig-drm6_1

color_scheme_set("mix-red-blue")

viz_pars <-
  c("tau[1]",
    "alpha[1]",
    "phi[1]")

mcmc_combo(drm_6$draws(variables = viz_pars),
           combo = c("trace", "dens_overlay"),
           facet_args = list(labeller = label_parsed))
```


@fig-drm6_2 shows the traceplots (left) and overlay-ed (right) density plots for
the regression coefficients associated with @eq-lmrec.

```{r}
#| label: fig-drm6_2
#| fig-cap: "Regression coefficients for recruitment."

mcmc_combo(drm_6$draws(variables = "coef_r"),
           combo = c("trace", "dens_overlay"))
```


@fig-drm6_3 shows the traceplots (left) and overlay-ed (right) density plots for
the regression coefficients associated with @eq-lmsurv.

```{r}
#| label: fig-drm6_3
#| fig-cap: "Regression coefficients for survival."

mcmc_combo(drm_6$draws(variables = "coef_m"),
           combo = c("trace", "dens_overlay"))
```


@fig-drm6_4 shows the traceplots (left) and overlay-ed (right) density plots for
the regression coefficients associated with @eq-linpreds.

```{r}
#| label: fig-drm6_4
#| fig-cap: "Regression coefficients for the probability of observing 0 density."
#| fig-height: 9

mcmc_combo(drm_6$draws(variables = "coef_t"),
           combo = c("trace", "dens_overlay"))
```


Despite the small number of posterior samples, the evidence for the graphs
suggests the chains have converged.


@fig-drm6_pp displays a posterior predictive model assessment
[@gabry2019visualization] for the DRM. The darker blue density is estimated from
the data directly, while the lighter blue densities are estimated from the
posterior predictive distribution (they can be seen as posterior samples from
"fitted" values). The similarity among the darker and lighter curves indicate
appropriate fit.

```{r}
#| label: fig-drm6_pp
#| fig-cap: "Posterior predictive check."

y_pp <-
  drm_6$draws(variables = "y_pp",
              format = "matrix")

ppc_dens_overlay(data$y,
                 y_pp[sample(seq_len(NROW(y_pp)),
                             size = 200), ])
```

### SDM

@fig-sdm_1 depicts the traceplots (left) and overlay-ed (right) density plots
for the parameter $\phi$ and the two parameters (`sigma_r` and $\rho$)
associated to the AR(1) random effect used to introduce temporal dependence to
the density. In general, the chains seem to mix well and the estimated
posterior densities associated to different chains are fairly similar, which is
a good indication of convergence.

```{r}
#| label: fig-sdm_1

viz_pars <-
  c("tau[1]",
    "alpha[1]",
    "phi[1]")

mcmc_combo(sdm_1$draws(variables = viz_pars),
           combo = c("trace", "dens_overlay"),
           facet_args = list(labeller = label_parsed))
```


@fig-sdm_2 shows the traceplots (left) and overlay-ed (right) density plots for
the regression coefficients associated with the second part of @eq-linpreds.

```{r}
#| label: fig-sdm_2
#| fig-cap: "Regression coefficients for positive density."

mcmc_combo(sdm_1$draws(variables = "coef_r"),
           combo = c("trace", "dens_overlay"))
```


@fig-drm6_4 shows the traceplots (left) and overlay-ed (right) density plots for
the regression coefficients associated with the first part of @eq-linpreds.

```{r}
#| label: fig-sdm_3
#| fig-cap: "Regression coefficients for the probability of observing 0 density."
#| fig-height: 9

mcmc_combo(sdm_1$draws(variables = "coef_t"),
           combo = c("trace", "dens_overlay"))
```


Despite the small number of posterior samples, the evidence for the graphs
suggests the chains have converged.


@fig-sdm_pp displays a posterior predictive model assessment
[@gabry2019visualization] for the SDM. The darker blue density is estimated from
the data directly, while the lighter blue densities are estimated from the
posterior predictive distribution (they can be seen as posterior samples from
"fitted" values). The similarity among the darker and lighter curves indicate
appropriate fit.

```{r}
#| label: fig-sdm_pp
#| fig-cap: "Posterior predictive check."

y_pp <-
  sdm_1$draws(variables = "y_pp",
              format = "matrix")

ppc_dens_overlay(data_sdm$y,
                 y_pp[sample(seq_len(NROW(y_pp)),
                             size = 200), ])
```


### Forecasting


The code below computes the forecasts for the best DRM and SDM.

```{r}
#| label: forecast-1

x_tt <- model.matrix(form_t,
                     data = dat_test)
x_mt <- model.matrix(form_m,
                     data = dat_test)
x_rt <- model.matrix(form_r,
                     data = dat_test)

x_mpast <-
  model.matrix(form_m,
               data = filter(dat_train, year == max(year)))

forecast_drm_6 <- predict_drm(drm = drm_6,
                              drm_data = data,
                              ntime_for =
                                length(unique(dat_test$year)),
                              x_tt = x_tt,
                              x_rt = x_rt,
                              x_mt = x_mt,
                              x_mpast = x_mpast,
                              f_test = f_test[, -1],
                              seed = 125,
                              cores = 4)
##--- * SDM ----

fitted_params <-
  sdm_1$draws(variables = c("coef_t", "coef_r",
                            "z_t",
                            "alpha", "tau",
                            "phi"))

forecast_data2 <-
  list(n_patches = data_sdm$n_patches,
       n_time = length(unique(dat_test$year)),
       time = dat_test$time - min(dat_test$time) + 1,
       n_ages = 1,
       n_time_train = data_sdm$n_time,
       time_ar = data_sdm$time_ar,
       cloglog = data_sdm$cloglog,
       likelihood = data_sdm$likelihood,
       K_z = data_sdm$K_z,
       Z = x_tt,
       K_x = data_sdm$K_x,
       X = x_rt)

forecast_data2$N <- forecast_data2$n_patches * forecast_data2$n_time

forecast_sdm_1 <-
  forecast_sdm$
  generate_quantities(fitted_params = fitted_params,
                      data = forecast_data2,
                      seed = 125,
                      parallel_chains = 4)
```

### Forecast $\mu_{t, i}$

@fig-forecast_mu displays the forecasted expected densities conditioned on
encountering (i.e., $Y_{t, i} > 0$).

```{r}
#| label: fig-forecast_mu
#| fig-cap: Forecast DRM.

mu_drm <-
  forecast_drm_6$draws(variables = "mu_proj",
                     format = "draws_df") |>
  tidyr::pivot_longer(cols = starts_with("mu_proj"),
                      names_to = "pair",
                      values_to = "expected") |>
  group_by(pair) |>
  summarise(ll = quantile(expected, probs = .05),
            l = quantile(expected, probs = .1),
            m = median(expected),
            u = quantile(expected, probs = .9),
            uu = quantile(expected, probs = .95)) |>
  ungroup() |>
  mutate(pair = gsub("\\D", "", pair)) |>
  mutate(pair = as.integer(pair)) |>
  arrange(pair)

mu_drm <-
  dat_test |>
  select(dens, lat_floor, patch, year) |>
  mutate(pair = row_number()) |>
  left_join(mu_drm, by = "pair") |>
  mutate(model = "DRM")

mu_sdm <-
  forecast_sdm_1$draws(variables = "mu_proj",
                     format = "draws_df") |>
  tidyr::pivot_longer(cols = starts_with("mu_proj"),
                      names_to = "pair",
                      values_to = "expected") |>
  group_by(pair) |>
  summarise(ll = quantile(expected, probs = .05),
            l = quantile(expected, probs = .1),
            m = median(expected),
            u = quantile(expected, probs = .9),
            uu = quantile(expected, probs = .95)) |>
  ungroup() |>
  mutate(pair = gsub("\\D", "", pair)) |>
  mutate(pair = as.integer(pair)) |>
  arrange(pair)

mu_sdm <-
  dat_test |>
  select(dens, lat_floor, patch, year) |>
  mutate(pair = row_number()) |>
  left_join(mu_sdm, by = "pair") |>
  mutate(model = "SDM")

mu_proj <- bind_rows(mu_drm, mu_sdm)

my_color <- "#0465cf"
ggplot(data = mu_proj,
         aes(x = year,
             y = m)) +
  geom_ribbon(aes(ymin = ll, ymax = uu),
              fill = my_color,
              alpha = .4) +
  geom_ribbon(aes(ymin = l, ymax = u),
              fill = my_color,
              alpha = .4) +
  geom_line() +
  geom_point() +
  theme_bw() +
  facet_grid(model ~ patch,
             scales = "free_y") +
  labs(y = expression(mu),
       x = "Year")
```


### Forecast $Y_{t, i}$


@fig-forecast_y displayes the forecasts for the response variable through its
posterior predictive distribution. That is, unlike @fig-forecast_mu, it accounts
for the zero-augmentation.

```{r}
#| label: fig-forecast_y

y_drm <-
  forecast_drm_6$draws(variables = "y_proj",
                       format = "draws_df") |>
  tidyr::pivot_longer(cols = starts_with("y_proj"),
                      names_to = "pair",
                      values_to = "expected") |>
  group_by(pair) |>
  summarise(ll = quantile(expected, probs = .05),
            l = quantile(expected, probs = .1),
            m = median(expected),
            u = quantile(expected, probs = .9),
            uu = quantile(expected, probs = .95)) |>
  ungroup() |>
  mutate(pair = gsub("\\D", "", pair)) |>
  mutate(pair = as.integer(pair)) |>
  arrange(pair)

y_drm <-
  dat_test |>
  select(dens, lat_floor, patch, year) |>
  mutate(pair = row_number()) |>
  left_join(y_drm, by = "pair") |>
  mutate(model = "DRM")

y_sdm <-
  forecast_sdm_1$draws(variables = "y_proj",
                       format = "draws_df") |>
  tidyr::pivot_longer(cols = starts_with("y_proj"),
                      names_to = "pair",
                      values_to = "expected") |>
  group_by(pair) |>
  summarise(ll = quantile(expected, probs = .05),
            l = quantile(expected, probs = .1),
            m = median(expected),
            u = quantile(expected, probs = .9),
            uu = quantile(expected, probs = .95)) |>
  ungroup() |>
  mutate(pair = gsub("\\D", "", pair)) |>
  mutate(pair = as.integer(pair)) |>
  arrange(pair)

y_sdm <-
  dat_test |>
  select(dens, lat_floor, patch, year) |>
  mutate(pair = row_number()) |>
  left_join(y_sdm, by = "pair") |>
  mutate(model = "SDM")

y_proj <- bind_rows(y_drm, y_sdm)

ggplot(data = y_proj,
         aes(x = year,
             y = patch,
             fill = m)) +
  geom_tile() +
  scale_fill_viridis_c(option = "H") +
  scale_y_continuous(breaks = 1:10) +
  facet_wrap(~ model) +
  theme_bw()

my_color <- "#0465cf"
ggplot(data = y_proj,
         aes(x = year,
             y = m)) +
  geom_ribbon(aes(ymin = ll, ymax = uu),
              fill = my_color,
              alpha = .4) +
  geom_ribbon(aes(ymin = l, ymax = u),
              fill = my_color,
              alpha = .4) +
  geom_line() +
  geom_point() +
  geom_point(aes(x = year, y = dens),
             inherit.aes = FALSE,
             color = 2) +
  facet_wrap(patch ~ model,
             scales = "free_y") +
  theme_bw() +
  facet_grid(model ~ patch,
             scales = "free_y") +
  labs(y = expression(mu),
       x = "Year")
```

@tbl-assessment suggests the DRM delivers more reliable forecasts. However, the
coverage of the prediction intervals (PIC) are above the nominal level for the
DRM, while the ones from the SDM are below the nominal level.

```{r}
#| label: tbl-assessment
#| tbl-cap: Forecasting assessment.

int_score <- function(y, l, u, alpha) {
  alpha <- 2 / alpha
  ind_l <- as.numeric(y < l)
  ind_u <- as.numeric(y > u)
  (u - l) + alpha * ((l - y) * ind_l + (y - u) * ind_u)
}

y_proj |>
  mutate(bias = dens - m) |>
  mutate(rmse = bias * bias) |>
  mutate(abias = abs(bias)) |>
  mutate(is1 = int_score(dens, l = ll, u = uu, alpha = .1)) |>
  mutate(is2 = int_score(dens, l = l, u = u, alpha = .2)) |>
  mutate(cvg1 = data.table::between(x = dens, lower = ll, upper = uu)) |>
  mutate(cvg2 = data.table::between(x = dens, lower = l, upper = u)) |>
  group_by(model) |>
  summarise(across(bias:cvg2, mean)) |>
  ungroup() |>
  rename_all(toupper) |>
  rename("Model" = "MODEL",
         "IS (90%)" = "IS1",
         "IS (80%)" = "IS2",
         "PIC (90%)" = "CVG1",
         "PIC (80%)" = "CVG2")
```


### Centroids

@fig-centroids depicts the centroids estimated by the DRM and SDM on the left
and right-side of the panel, respectively.


```{r}
#| label: fig-centroids
#| fig-cap: Density weighted centroids estimated by the DRM and SDM. The red vertical line indicates where the "forecast" period starts.

centroid_drm <-
  centroids_compiled$
  generate_quantities(fitted_params =
                        posterior::bind_draws(
                                       drm_6$draws(variables = "y_pp"),
                                       forecast_drm_6$draws(variables = "y_proj")
                                       ),
                      data = list(
                          N1 = nrow(dat_train),
                          N2 = nrow(dat_test),
                          n_c = 1,
                          coords =
                            matrix(c(dat_train$lat_floor,
                                     dat_test$lat_floor),
                                   ncol = 1),
                          n_t = length(unique(c(dat_train$time,
                                                dat_test$time)))
                      ),
                      seed = 2025,
                      parallel_chains = 4)

centroid_sdm <-
  centroids_compiled$
  generate_quantities(fitted_params =
                        posterior::bind_draws(
                                       sdm_1$draws(variables = "y_pp"),
                                       forecast_sdm_1$draws(variables = "y_proj")
                                       ),
                      data = list(
                          N1 = nrow(dat_train),
                          N2 = nrow(dat_test),
                          n_c = 1,
                          coords =
                            matrix(c(dat_train$lat_floor,
                                     dat_test$lat_floor),
                                   ncol = 1),
                          n_t = length(unique(c(dat_train$time,
                                                dat_test$time)))
                      ),
                      seed = 2025,
                      parallel_chains = 4)


centroid_all <-
  bind_rows(
      centroid_drm$draws() |>
      bayesplot::mcmc_intervals_data() |>
      mutate(type = gsub("[[:digit:][:punct:]]", "", parameter),
             .before = parameter) |>
      mutate(parameter = gsub("\\D", "", parameter)) |>
      mutate(parameter = ifelse(nchar(parameter) == 2,
                                substr(parameter, 1, 1),
                                substr(parameter, 1, 2))) |>
      mutate(year = as.integer(parameter) + min_year,
             .before = parameter,
             model = "DRM"),
      centroid_sdm$draws() |>
      bayesplot::mcmc_intervals_data() |>
      mutate(type = gsub("[[:digit:][:punct:]]", "", parameter),
             .before = parameter) |>
      mutate(parameter = gsub("\\D", "", parameter)) |>
      mutate(parameter = ifelse(nchar(parameter) == 2,
                                substr(parameter, 1, 1),
                                substr(parameter, 1, 2))) |>
      mutate(year = as.integer(parameter) + min_year,
             .before = parameter,
             model = "SDM")
  )

ggplot(data = filter(centroid_all,
                     type == "centroid"),
         aes(x = year,
             y = m)) +
  geom_ribbon(aes(ymin = ll, ymax = hh),
              fill = my_color,
              alpha = .4) +
  geom_ribbon(aes(ymin = l, ymax = h),
              fill = my_color,
              alpha = .4) +
  geom_line() +
  geom_point() +
  geom_vline(xintercept = first_year_forecast,
             col = 2, lty = 2,
             lwd = 1.2) +
  facet_wrap( ~ model) +
  theme_bw()
```
