---
title: "Get Started"
author: Lucas da Cunha Godoy
bibliography: [notes.bib, zim.bib]
execute:
  eval: true
vignette: >
  %\VignetteIndexEntry{Get Started}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup

library(drmr)
library(sf) ## "mapping"
library(ggplot2) ## graphs
library(bayesplot) ## and more graphs
library(dplyr)
```

We will use a dataset included with the package. In particular, this data is
similar to the one analyzed in @fredston2025dynamic. To load the data, we run:

```{r}
#| label: data
data(sum_fl)
```

## Data processing & splitting

To assess model predictions, the chunk below splits the data into train and
test.

```{r}
#| label: split
## 5 years-ahead predictions
first_year_forecast <- max(sum_fl$year) - 5

first_id_forecast <-
  first_year_forecast - min(sum_fl$year) + 1

years_all <- length(unique(sum_fl$year))
years_train <- years_all[years_all < first_id_forecast]
years_test <- years_all[years_all >= first_id_forecast]

## splitting data
dat_test <- sum_fl |>
  filter(year >= first_year_forecast)

dat_train <- sum_fl |>
  filter(year < first_year_forecast)
```

Finally, I transform our response variable into a density.

```{r}
#| label: ytransform
dat_train <- dat_train |>
  mutate(dens = y / area_km2,
         .before = y)

dat_test <- dat_test |>
  mutate(dens = y / area_km2,
         .before = y)
```

The elements we need to fit a DRM are:

* A response variable representing a density

* A variable encoding time-points

* A variable encoding patches (or sites).

The current version of the functions assumes that we have rows in the dataset
for every combinations of "time" and "site". That is, if a dataset comprises
observations around the density of a given species across 20 years and 10
patches, we expect to have 20 rows in the dataset per patch. Equivalently, we
would expect 10 rows in the dataset per year.


The first and last 10 rows of the `dat_train` object look as follows:

```{r}
#| label: data-format

head(dat_train[, c("year", "patch",  "y", "stemp")],
     n = 10)
tail(dat_train[, c("year", "patch",  "y", "stemp")],
     n = 10)
```

`stemp` is an explanatory variable denoting the sea surface temperature (SST).


## Fitting & diagnosing a model

Provided the dataset is properly formatted, a simple model considering the
existance of `8` age groups and a natural mortality rate of `0.25` can be fit as
follows:

```{r}
#| label: fit_drm
#| eval: true
my_drm <-
  fit_drm(.data = dat_train,
          y_col = "dens",
          time_col = "year",
          site_col = "patch",
          n_ages = 8,
          m = 0.25,
          seed = 2025)
```


The `my_drm` is an object of class `drmmodels`. We have specific methods
(i.e., specialized functions) for this object. For example, one can get
parameter estimates and some convergence diagnostics using the `summary` method:
```{r}
#| label: summary_drm

summary(my_drm)
```

In the output above, the column `variable` represents distinguish the model
parameters (for the details on the model parameters see the
[Theory](pinskylab.github.io/drmr/theory.html) vignette), while `mean` and `sd`,
respectively, are the mean and standard deviation of the respective posterior
distributions. The `rhat` and `ess_bulk` are diagnostic statistics. In general,
an `rhat < 1.01` suggests convergence, but one should also look into traceplots
to further assess convergence. The `ess_bulk` is the effective sample size in
the bulk of the posterior. The columns starting with `q` represent percentiles
of the posterior distribution.



We can use our `draws` method along with the functions from the `bayesplot`
package to visualize the traceplots and estimated densities of the posterior
distributions:
```{r}
#| label: mcmc_combo
draws(my_drm) |>
  mcmc_combo(combo = c("trace", "dens_overlay"),
             facet_args = list(labeller = label_parsed))
```

## Updating the model

If we want to update a few arguments of our `fit_drm` call, we can resort to the
`update` method. For example, suppose we want to make the probability of absence
depend on a measure of "effort". In our case, number of hauls `n_hauls` can be
used as a proxy for effort. In addition, let us establish a non-linear
relationship between recruitment and SST (the `stemp` column in our data). We
```{r}
#| label: update_method
#| eval: false

my_drm2 <-
  update(my_drm,
         formula_zero = ~ 1 + n_hauls,
         formula_rec = ~ 1 + stemp + I(stemp * stemp), ## establising a
                                                       ## quadratic relationship
                                                       ## between recruitment and stemp
         .toggles = list(ar_re = "rec"), ## adding an autoregressive term to recruitment
         algo_args = list(parallel_chains = 4)) ## the `algo_args` is a list
                                                ## that allows us to modify the
                                                ## parameters of the MCMC
                                                ## algorithm
```
